---
title: STA5092Z 2026 Assignment 1
author:
  - name: Pamela Afful
    email: affpam001@myuct.ac.za
    # affiliations: 
    #     - id: uct
    #       name: University of Cape Town
    #       department: Statistical Sciences Department
    #       address: Rondebosch
    #       city: Cape Town
    #       state: Western Cape
    #       postal-code: 7701
          
abstract: | 
   I will doing some EDA on very messy data, yay!
   
keywords: 
  - pied babbler *comeback and edit properly
#date: last-modified
#bibliography: eda.bib
format:
  pdf
    # code-line-numbers: true
    # standalone: true
    # toc: true
    # number-sections: true
    # engine: xelatex
    # text: |
    #     \usepackage{lmodern}
    #     \usepackage[T1]{fontenc}
    #     \usepackage{amsmath}
    #     \usepackage{bm}
    # keep-tex: true
    # journal:
    #   name: Journal Name
    #   cite-style: authoryear
execute: 
  cache: true
  warning: false
  message: false
  echo: false
---

```{r}

knitr::opts_chunk$set(dev = "pdf")
```

```{r}
# Import packages
#| echo: false
library(lubridate)
library(dplyr)
library(janitor)
library(tidyverse)
library(scales)
library(ggridges)
```

# Part 1- Data Wrangling

### Load Data

2\) Load data and transform into tibble objects

```{r data_load}
#| echo: false
fledg_survival=tibble::as_tibble(read.csv('Data/Data.Bourne_FledglingSurvival.csv'))
fledg_survival_meta=tibble::as_tibble(read.csv('Data/Data.Bourne_FledglingSurvival_Meta.csv'))

nest_all=tibble::as_tibble(read.csv('Data/Data.Bourne_NestlingsAllData.csv'))
nest_all_meta=tibble::as_tibble(read.csv('Data/Data.Bourne_NestlingsAllData_Meta.csv'))

nest_succ=tibble::as_tibble(read.csv('Data/Data.Bourne_NestSuccess.csv'))
nest_succ_meta=tibble::as_tibble(read.csv('Data/Data.Bourne_NestSuccess_Meta.csv'))

# View data

nest_all %>% select(Date)

fledg_survival %>% select(Date)
```

### Merge Tibble Dataframes

3\)

### DataFrame Column Data Types

#### Key Findings:

**fledg_survival** dataframe

-   Data types in columns are mostly split between character and int/numeric data types

-   **Date** column in fledg_survival is stored as a character- will need to be converted. The data format is of the format dd-mm-yy

-   **SurvMassAtInd** column contains null values

**nest_all** dataframe

-   In addition to numeric and int/numeric data types, columns also include logical/boolean data types

-   **Date** is also a character data type, similar to the format in the **fledg_survival** dataframe

Merging steps:

1.  First, convert date columns in both data frames into datetime columns using the lubridate ymd function

2.  Find common column names in both the **nest_all** and **fledg_survival** data sets to find columns to join on

3.  Check and ensure that the datatypes of these common columns are the same in both data sets- using the janitor package- and change any datatype mismatches to the same data type

4.  Both GrpSizeAD and GrpSizeTotal columns are stored as characters in the nest_all data frame, as they both contain #NA values

5.  These will replaced with na values instead, once they are coerced into integer values

```{r data_join}
#| echo: false

# Find common columns 
common_cols=intersect(names(fledg_survival), names(nest_all))

# Select the common cols in both data frames
fledg_survival_common =fledg_survival %>% select( all_of(common_cols))

nest_all_common =nest_all %>% select( all_of(common_cols))

# compare data types between dataframes
mismatches =compare_df_cols(fledg_survival_common, nest_all_common, return = "mismatch")


#columns, GrpSizeAD and GrpSizeTotal are different data types 

# In particular, GrpSizeAD GrpSizeTotal in nest_all and  contains #NA values- so these are bieng read in as charaters

distinct_grp_size_tot=nest_all %>% distinct(GrpSizeTotal)
distinct_grp_size_grp_sizead=nest_all %>% distinct(GrpSizeAD)


#  GrpSizeAD and GrpSizeTotal in nest_all df to the ints
# convert date into ymd
# Coercing #NA values into integers will forces those values to be R nas, which are much easier to work with 


nest_all_clean=nest_all %>% mutate(Date=dmy(Date),GrpSizeAD=as.integer(GrpSizeAD),GrpSizeTotal=as.integer(GrpSizeTotal) )

fledg_survival_clean=fledg_survival %>% mutate(Date=dmy(Date))

# Merge the two

piedbabbler=left_join(nest_all_clean,fledg_survival_clean, by=common_cols)


```

# Part 2- Data Summaries and Visualizations

1\) **piedbabbler** samplesize

```{r row_count_check}
#| echo: false
# number of rows
all_birds=piedbabbler %>% distinct(BirdID) %>% pull(BirdID)

fl_birds=fledg_survival %>% distinct(BirdID) %>% pull(BirdID)

diff_=setdiff(all_birds, fl_birds)

piedbabbler %>% filter(BirdID %in% diff_) %>% distinct(	
SurvInd)

piedbabbler %>% filter(FledgeDate=='') %>% select(FledgeDate)
```

2\)

#### Null Value Check

```{r null_val_check}

#| echo: false

check_cols=c('GrpSizeAD','GrpSizeTotal','BroodSize','Mass','TMaxMeas','No.ChicksFledge','Rain90','Drought')

# Count of null values per column
# Show as percentage
piedbabbler %>% select (all_of(check_cols)) %>% 
  mutate(across(all_of(check_cols), ~is.na(.))) %>% 
  summarise(across(everything(), ~sum(.))) 

  

```

35% of values i

```         
```

#### Outlier Analysis

Outlier Definition: An outlier is an individual point of data that is distant from other points in the data set. It is an anomaly in the dataset that may be caused by a range of errors in capturing, processing or manipulating data. \* comeback and reference

```{r out_lier}
#| echo: false

# An outlier is an individual point of data that is distant from other points in the dataset. It is an anomaly in the dataset that may be caused by a range of errors in capturing, processing or manipulating data. 

# https://www.seldon.io/outlier-detection-and-analysis-methods/

# For now, use box plots to visually identify cols with outliers

piedbabbler %>% select (all_of(check_cols)) %>% 
  pivot_longer(cols=check_cols,
               names_to='Column_Variable',
               values_to = 'value') %>% 
  ggplot(aes(y = value)) +
  geom_boxplot() +
  facet_wrap(~ Column_Variable, scales = "free_y") +
  labs(title = "Distribution of Features ",
       y = "values") +
  theme_minimal()
  


```

From their respective boxplots, the following columns seem to contain outliers, relative to their own distributions:

-   TMaxMeas

-   Mass

-   BroodSize

-   GrpSizeTotal

3\)

#### Mass and Date per Nest: Longitudinal Dataframe

```{r mass_per_nest_id}

#| echo: false

# Create via groupby
# Comeback and check tomorrow

piedbabbler %>% group_by(NestID,Date) %>% 
  summarize(total_mass=sum(Mass))
```

4\)

TMaxMeas has 35% of its values as nulls. As such, the aggregate mean Mass value per bucketed TMaxMeas will be use to adjust for this

```{r crit_tem}

# Come back and fix buckets
#| echo: false
piedbabbler %>% select(Mass,TMaxMeas) %>%
    drop_na() %>% 
mutate(TMaxMeas_bin = cut(TMaxMeas, seq( min(TMaxMeas), max(TMaxMeas) + 3, 3), right = FALSE)) %>%
group_by(TMaxMeas_bin) %>%
summarise(mean_mass=mean(Mass)) %>%
  ggplot(aes(x=TMaxMeas_bin,y = mean_mass,group=1)) +
  geom_line()+
  geom_vline(xintercept = '[21.5,24.5)', color = "blue", linetype = "dashed", size = 1.5)+
  
  annotate("text", 
           x = '[24.5,27.5)',  # Adjust x for horizontal positioning relative to the line
           y = 50,              # Adjust y for vertical positioning (based on plot's y-scale)
           label = paste("Tcritical"), 
           color = "blue",
           angle = 0)+
  labs(title = "Average Chick Mass by TMaxMeas ",
       x='TMaxMeas bucket',y = "average chick mass") +
  theme_minimal()

  


```

5\)

Mass over time per nest per season

Since we have way too distinct NestIDs, the following grouping approach will be used:\

1.  First get the size of each nest, defined by the count of birds in each nest- since this value, changes over time, the starting nest size for each group will be used

2.  Use this count as the grouping variable to compare mass over time

3.  Show the individual nestID groups for the "best" performing birds

4.  Also, since there are not enough observations for each nest ID, per year, the year variable will be normalized relatve to the start date/year of the nest

```{r plt_c_g}
# Split this code over multiple cells
library(ggplot2)
library(plotly)
factor_ord=sapply(c(1,2,3,4,5),as.factor)

season_groups=c(1,2,3)
seasons_val=piedbabbler %>% distinct(Season) %>% 
  arrange() %>% 
  mutate(num_index=1:n_distinct(Season)) %>% 
  mutate(num_index_break=cut(num_index,breaks=3,labels=c(1,2,3))) 
seasons_val

nest_sizes=piedbabbler %>% group_by(NestID) %>% 
  summarize(nest_size=n_distinct(BirdID))

piedbabbler_2=left_join(seasons_val,piedbabbler,by='Season')
piedbabbler_2=left_join(piedbabbler_2,nest_sizes,by='NestID')
piedbabbler_2

distinct_sizes=c(1, 2,  3,  4,  5)

fig_1=piedbabbler_2 %>% filter(num_index_break==1) %>% 
  group_by(Date,Season,nest_size) %>%
  summarize(total_nest_mass=mean(Mass)) %>%
  mutate(nest_size=factor(nest_size,levels=distinct_sizes)) %>% 
  arrange(nest_size) %>% 
  group_by(nest_size,Season) %>% 
  mutate(obs_id = factor(row_number())) %>% 
  ggplot(aes(x=obs_id,y =total_nest_mass,group = 1,color=nest_size)) +
  geom_point() +
  geom_line(aes(group = nest_size))+
  facet_wrap(~Season,scales='free',ncol=2) +
  labs(title = "Chick Mass Over Time Per Nest ID: Seasons C to G ",
       x='Observation ID',y = "Chick Mass") +
  theme_minimal()
fig_1


```

```{r plt_h_j}

distinct_sizes=c(1, 2,  3,  4,  5)
fig_2=piedbabbler_2 %>% filter(num_index_break==2) %>% 
  group_by(Date,Season,nest_size) %>%
  summarize(total_nest_mass=mean(Mass)) %>%
  mutate(nest_size=factor(nest_size,levels=distinct_sizes)) %>% 
  arrange(nest_size) %>% 
  group_by(nest_size,Season) %>% 
  mutate(obs_id = factor(row_number())) %>% 
  ggplot(aes(x=obs_id,y =total_nest_mass,group = 1,color=nest_size)) +
  geom_point() +
  geom_line(aes(group = nest_size))+
  facet_wrap(~Season,scales='free',ncol=2) +
  labs(title = "Chick Mass Over Time Per Nest ID: Seasons H to L",
       x='Observation ID',y = "Average Chick Mass") +
  theme_minimal()
fig_2


```

```{r plt_m_r}
distinct_sizes=c(1, 2,  3,  4,  5)
fig_3=piedbabbler_2 %>% filter(num_index_break==3) %>% 
  group_by(Date,Season,nest_size) %>%
  summarize(total_nest_mass=mean(Mass)) %>%
  mutate(nest_size=factor(nest_size,levels=distinct_sizes)) %>% 
  arrange(nest_size) %>% 
  group_by(nest_size,Season) %>% 
  mutate(obs_id = factor(row_number())) %>% 
  ggplot(aes(x=obs_id,y =total_nest_mass,group = 1,color=nest_size)) +
  geom_point() +
  geom_line(aes(group = nest_size))+
  facet_wrap(~Season,scales='free',ncol=2) +
  labs(title = "Chick Mass Over Time Per Nest ID: Seasons M to R",
       x='Observation ID',y = "Chick Mass") +
  theme_minimal()
fig_3
```

Although the time ranges are inconsistent and disparate across seasons, nest sizes with 3 chicks, show higher masses over time. This will be used as the criterion to filter and observe individual nest ids that fall in this category.

```{r mess_ev}
fig_4=piedbabbler_2 %>% 
  filter(nest_size==3) %>% 
  group_by(Date,NestID,Season) %>% 
  summarize(total_nest_mass=sum(Mass)) %>% 
  ggplot(aes(x=Date,y =total_nest_mass,group = 1,color=NestID)) +
  geom_point() +
  geom_line(aes(group = NestID))+
  facet_wrap(~Season,scales='free',ncol=2) +
  labs(title = "Chick Mass Over Time Per Nest ID: Seasons M to R",
       x='date',y = "Chick Mass") +
  theme_minimal()
fig_4




```

6\)

**Yearly Average Mass, Faceted by Season**

```{r}
str(piedbabbler)
```

```{r chick_mass_drought}
#| echo: false
piedbabbler %>% mutate(dt_year=year(Date)) %>% 
  select(dt_year,Mass,BroodSize,GrpSizeAD,No.ChicksFledge,Rain90,Season,Drought) %>% group_by(dt_year, Drought) %>% #,Drought Season
  summarize(avg_mass=mean(Mass),avg_brsize=mean(BroodSize),avg_grp_size=mean(GrpSizeAD), avg_chick_fledge=mean(No.ChicksFledge), avg_rain=mean(Rain90)) %>% 
  mutate(Drought=as.character(Drought) ) %>% 
  drop_na(Drought)%>% 
       ggplot(aes(x=dt_year,y = avg_mass)) +
  geom_line() + #geom_c(stat='identity') 
  facet_wrap(~Drought , scales = "free_y" ) +
  labs(title = "Chick Mass Over Time Per Nest ID ",
       x='date',y = "Average Chick Mass") +
  theme_minimal()


  
```

**Average Bird Size Over Time- Faceted By Season**

```{r chick_size_drought}
#| echo: false

season_labels <- c(
  `1` = "Drought",
  `0` = "No Drought"
)
piedbabbler %>% mutate(dt_year=year(Date)) %>% 
  select(dt_year,Mass,BroodSize,GrpSizeAD,No.ChicksFledge,Rain90,Season,Drought) %>% group_by(dt_year, Drought) %>% #,Drought Season
  summarize(avg_mass=mean(Mass),avg_brsize=mean(BroodSize),avg_grp_size=mean(GrpSizeAD), avg_chick_fledge=mean(No.ChicksFledge), avg_rain=mean(Rain90)) %>% 
  mutate(Drought=as.character(Drought) ) %>% 
  drop_na(Drought)%>% 
       ggplot(aes(x=dt_year,y = avg_brsize)) +
  geom_line() + #geom_c(stat='identity') 
  facet_wrap(~Drought , scales = "free_y",labeller = as_labeller(season_labels) ) +
  labs(title = "Average Bird Size Over Time Faceted by Season ",
       x='date',y = "Average Bird Size") +
  theme_minimal()
```

```{r grp_size_drought}
#| echo: false


season_labels <- c(
  `1` = "Drought",
  `0` = "No Drought"
)
piedbabbler %>% mutate(dt_year=year(Date)) %>% 
  select(dt_year,Mass,BroodSize,GrpSizeTotal,No.ChicksFledge,Rain90,Season,Drought) %>% group_by(dt_year, Drought) %>% #,Drought Season
  summarize(avg_mass=mean(Mass),avg_brsize=mean(BroodSize),avg_grp_size=mean(GrpSizeTotal), avg_chick_fledge=mean(No.ChicksFledge), avg_rain=mean(Rain90)) %>% 
  mutate(Drought=as.character(Drought) ) %>% 
  drop_na(Drought)%>% 
       ggplot(aes(x=dt_year,y = avg_grp_size)) +
  geom_line() + #geom_c(stat='identity') 
  facet_wrap(~Drought , scales = "free_y" ,labeller = as_labeller(season_labels)) +
  labs(title = "Average Group Size Over Time Faceted by Season ",
       x='date',y = "Average Group Size") +
  theme_minimal()
```

```{r chick_fledge_drought}
#| echo: false


season_labels <- c(
  `1` = "Drought",
  `0` = "No Drought"
)
piedbabbler %>% mutate(dt_year=year(Date)) %>% drop_na(No.ChicksFledge) %>% 
  select(dt_year,No.ChicksFledge,Rain90,Season,Drought) %>% 
  group_by(dt_year, Drought) %>% 
  summarize(avg_chick_fledge=mean(No.ChicksFledge)) %>%
  mutate(Drought=as.character(Drought) ) %>%
  drop_na(Drought)%>%
       ggplot(aes(x=dt_year,y = avg_chick_fledge)) +
  geom_line() + #geom_c(stat='identity')
  facet_wrap(~Drought , scales = "free_y" ,labeller = as_labeller(season_labels)) +
  labs(title = "Yearly Average Number of Fledging Chicks Over Time Faceted by Season ",
       x='year',y = "Average Number of Fledging Chicks") +
  theme_minimal()
```

```{r rain_drought}
piedbabbler %>% mutate(dt_year=year(Date)) %>%
  drop_na(Rain90) %>% 
drop_na(No.ChicksFledge) %>%
select(dt_year,Mass,TotRain,Season,Drought) %>%
group_by(dt_year, Drought) %>%
summarize(avg_rain=mean(TotRain)) %>%
mutate(Drought=as.character(Drought) ) %>%
drop_na(Drought)%>%
     ggplot(aes(x=dt_year,y = avg_rain)) +
geom_line() + #geom_c(stat='identity')
facet_wrap(~Drought , scales = "free_y" ,labeller = as_labeller(season_labels)) +
labs(title = "Yearly Average Rain Fall Faceted by Season ",
     x='year',y = "Average Rain ") +
theme_minimal()
```

6\) Rate of mass change per day per nesting

Steps for finding these suspicious entries

1.  First group Mass by Group and Date to get Mass per Nest over time
2.  Get lagged values of Date and Mass
3.  Calculate the percentage Mass change for each row per nest group
4.  Calculate the date difference, in days, between Date values and lagged date values
5.  Filter for date differences equal to 1 then check which growth rates are suspicious (\>50%)

```{r sus}

#| echo: false

# This data is not daily!!!!

piedbabbler  %>% 
  select(Date,Group,Mass) %>% 
  arrange(Group) %>% 
  group_by(Date,Group) %>% 
  summarize(total_mass=sum(Mass)) %>% 
  group_by(Group) %>% 
  arrange(Group) %>% 
  group_by(Group) %>% 
  mutate(lagged_mass=lag(total_mass,n=1), lagged_date=lag(Date,n=1)) %>%
 mutate(growth_rate=((total_mass-lagged_mass)/ lagged_mass)*100,date_diff=as.numeric(Date-lagged_date )) %>% 
  drop_na(growth_rate) %>% 
  filter(date_diff==1) %>% 
 distinct(Group)

  

  
```

All these above values are NAs\*, so I guess there was no data for these values(?)

8\)

```{r mass_fleg_ind}
piedbabbler %>% 
  select(Group,Mass,FledgeDate) %>% 
  mutate(FledgeDate=na_if(FledgeDate,'')) %>% 
  mutate(fledged=case_when(
    is.na(FledgeDate)~0,
    !is.na(FledgeDate)~1
  )) %>% 
  group_by(fledged,Group) %>% 
  summarize(tot_mass=sum(Mass)) %>%
  mutate(fledged=factor(fledged,levels=c(0,1))) %>% 
  ggplot(aes(y = tot_mass,color=fledged)) +
  geom_boxplot() +
  labs(title = "Distribution Mass By Fledging Indicator ",
       y = "total mass") +
  theme_minimal()
```

```{r mass_diff}
# Get Age btween Hatching and fleging
# Think of a way of finding the age where if a bird/nestling survives past, their liklihood for survival increases compared to birds that did not survive past that age
piedbabbler %>% select(NestID,HatchDate,FledgeDate,Date,Mass,AgeMeasD,	
SurvFledge) %>% 
  mutate(FledgeDate=na_if(FledgeDate,'')) %>% 
  rename(fledged=SurvFledge) %>% 
  group_by(fledged,AgeMeasD) %>% 
  summarise(avg_mass=mean(Mass)) %>% 
  mutate(fledged=factor(fledged,levels=c(0,1))) %>% 
  drop_na() %>% 
  pivot_wider(
    names_from=fledged,
    values_from=avg_mass
  ) %>% 
    rename('Fledged'='1','Not_Fledged'='0') %>% 
  mutate(mass_diff=Fledged-Not_Fledged) %>% 
  ggplot(aes(x=AgeMeasD,y = mass_diff)) +
  geom_col() +
  labs(title = "Average Hatchling Mass Difference By Hatchling Age- Factored By Flegding Status",x='Age(days)',
       y = "avg hatchling mass") +
  theme_minimal()


```

9\)

First check how the number of adult helpers are distributed across the data

```{r ad_help}
# Number distribution of adult heplers in the dataset
piedbabbler %>% select(GrpSizeAD) %>% 
  ggplot(aes(x = GrpSizeAD)) +
  geom_histogram()+
    labs(title = "Distribution of Ault Heplers",x='Adult Helper Count')+
  theme_minimal()
  

```

In this case, most nests had adult heplers of 3. Thus, to normalize the data, average fledgling count per adult helper count we'll be use to compare.

9\)

```{r help_cnt}
# We need to look at the incremental increase of adding more chicks

piedbabbler %>% drop_na(No.ChicksFledge) %>% 
  select(GrpSizeAD,No.ChicksFledge) %>% 
   arrange(desc(No.ChicksFledge)) %>% 
  group_by(GrpSizeAD) %>% 
  summarise(nr_chicks=mean(No.ChicksFledge)) %>% 
  mutate(nr_chicks_lag=lag(nr_chicks,n=1,default = first(nr_chicks))) %>% 
  mutate(change=nr_chicks-nr_chicks_lag,per_change=(nr_chicks-nr_chicks_lag)/nr_chicks_lag,label = percent(per_change, accuracy = 0.1)  ) %>% 
  ggplot(aes(x=GrpSizeAD,y = nr_chicks,label=per_change)) +
  geom_line() +
  geom_point() +
  geom_text(aes(label = label), vjust = -0.7) +
  labs(title = "Change in No. of Fledging Chicks by Adult Helper Count ",x='Adult Helper Count',y = "avg number of fleged chicks") +
  geom_vline(xintercept = 3, color = "blue", linetype = "dashed", size = 1.5)+
  
  annotate("text", 
           x = 4,  # Adjust x for horizontal positioning relative to the line
           y = 1,              # Adjust y for vertical positioning (based on plot's y-scale)
           label = paste("Max Change"), 
           color = "blue",
           angle = 0)+
 
  theme_minimal()








  




  

```

The average number fledged chicks sees an increase from 3 to 6, however it sees its most marked sustained decrease from 6 to 8 ( although this rebounds after 8)

```{r cum_per}

ch_sum_df=piedbabbler %>% drop_na(No.ChicksFledge) %>% 
  summarise(tot_chicks=sum(No.ChicksFledge))
ch_sum=ch_sum_df$tot_chicks

piedbabbler %>% drop_na(No.ChicksFledge) %>% 
  group_by(GrpSizeAD) %>% 
  summarize(nr_chicks=sum(No.ChicksFledge))  %>% 
   mutate(nr_chicks_cum_sum=cumsum(nr_chicks))  %>% 
  mutate(cum_percent=nr_chicks_cum_sum/ch_sum) %>% 
  mutate(cum_percent_change= cum_percent-lag(cum_percent,n=1,default = first(cum_percent)))  %>% 
  
    ggplot(aes(x=GrpSizeAD,y=cum_percent)) +
  geom_line() +
  geom_point() +
  geom_text(aes(label = round(cum_percent_change,2)  ), vjust = -0.7) +
  labs(title = "Change in No. of Fledging Chicks by Adult Helper Count ",x='Adult Helper Count',y = "cumulative percentage") +
  geom_vline(xintercept = 6, color = "blue", linetype = "dashed", size = 1.5)+
 
  annotate("text", 
            x = 6,  # Adjust x for horizontal positioning relative to the line
           y = 0.85,              # Adjust y for vertical positioning (based on plot's y-scale)
          label = paste("Max Change"), 
          color = "blue",
          angle = 0)+
 labs(title = "Cumulative Percentage Change in Chick Mass by Nr Adult Helpers",
      x='# Adult Helpers',y = "number of fledged chicks") +
  theme_minimal()
  





```

10\)

Comeback\*

Nestiling Mass Per Drought Per Nest Size

Define big nest size as GrpSizeTotal \>6

Small nest size GrpSizeTotal \<6

```{r drough_comp}

# Filter drought years

piedbabbler %>% filter(Drought==1) %>% 
 mutate(nest_size= case_when(GrpSizeTotal<6 ~'small',
                             GrpSizeTotal>6 ~'big') 
        ) %>% 
  select(Date,Mass,nest_size) %>% 
  group_by(Date,nest_size) %>% 
  summarize(avg_mass=mean(Mass)) %>% 
  group_by(nest_size) %>% 
  mutate(obs_id = row_number()) %>% 
  #filter(nest_size=='big') %>% 
    ggplot(aes(x=obs_id,y =avg_mass,color=nest_size)) +
  geom_line() +
  geom_point() +
  labs(title = "Mass over time facetted by Nest Size",x='Observation',y = "avg mass per nestling") +
  theme_minimal()




```

Smaller nests seem to survive longer over periods of drought\*

11\)

Seasonal Heatmap

We need to bucket the ambient values

Nesting mortality=number of dead birds/total number of birds

Come back and explain

```{r heat_map}
# calculate nestling mortality
surv=piedbabbler %>% select(Season, TMaxMeas,SurvInd,NestID) %>% 
  drop_na(TMaxMeas) %>% 
  mutate(TMaxMeas_bin = cut(TMaxMeas, seq( min(TMaxMeas), max(TMaxMeas) + 3, 3))) %>% group_by(TMaxMeas_bin,Season) %>% 
  summarize(survied_cnt=sum(SurvInd))

counts=piedbabbler %>% select(Season, TMaxMeas,NestID) %>% 
  drop_na(TMaxMeas) %>% 
  mutate(TMaxMeas_bin = cut(TMaxMeas, seq( min(TMaxMeas), max(TMaxMeas) + 3, 3))) %>% group_by(TMaxMeas_bin,Season) %>% 
  summarize(tot_count=length(NestID))

left_join(surv,counts, by=c('TMaxMeas_bin','Season')) %>% 
  mutate(mortality= 1-(survied_cnt/tot_count )     
           ) %>% 
  drop_na(TMaxMeas_bin) %>% 
  ggplot(aes(x = TMaxMeas_bin, y = Season, fill = mortality)) + # Map variables to aesthetics
  geom_tile(color = "white", lwd = 1.5, linetype = 1) + # Use geom_tile for the heatmap with borders
  coord_fixed() + # Ensure square tiles for a proper matrix look
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", # Customize the color scale
                       midpoint = 0.5, limit = c(0, 1), space = "Lab", 
                       name="Mortality Rate") +
  theme_minimal() + # Use a clean theme
  labs(title = "Heatmap of Mortality Per Season and Ambient Air Temprature",x='Ambient Air Bucket',y='Breeding Season') # Add a title
```

12\)

Ridge Plot

```{r ridge}
piedbabbler %>% 
  select(Mass,Season) %>% 
  ggplot(aes(x = Mass, y = Season)) +
  geom_density_ridges()+
  theme_minimal() +
  labs(title = "Distribution of Mass Per Season",x='Mass',y='Breeding Season') 
  

```

**Season Q** has the most bimodal distribution

13\) Comeback and think if a nice plot- Sankey Plot?

```{r}
 # Sankey Plot

library(ggplot2)
library(ggsankey)

# Example data (replace with your own data)
# You might need to use the make_long() function to format your data
data <- data.frame(
  Source = c("A", "A",'B1','B1','C1','c1'),
  Target = c("B1", "B2",'C1','C2','D1','D2'),
  Freq = c(5, 5,3,2,2,1)
)

# Transform for ggsankey
df <- data %>%
  make_long(Source, Target, value = Freq)

ggplot(df, aes(x = x, next_x = next_x, node = node, next_node = next_node, 
               fill = factor(node), label = node)) +
  geom_sankey(flow.alpha = 0.5, node.color = "black") +
  geom_sankey_label(size = 3, color = "black", fill = "white") +
  theme_sankey(base_size = 16) +
  labs(x = NULL) +
  theme(legend.position = "none")
```

```{r}
#  # Data Prep
# Bird ID is the most granular/unique identifier

# Stage 1: Total number of birds
# Total number of birds that hatched
# total number of birds that fleged?
# total number of birds that made it to nutritional independence
# Dotal number of birds that made it to 365 days
# totla number of birds that made it to dispersion

piedbabbler  %>% summarise(total_nr_birds=n_distinct(BirdID)) %>% 
  pull(total_nr_birds) # total number of birds 

# Total number of distinct birds
st_1=piedbabbler  %>% filter(!is.na(HatchDate)) %>% 
  summarise(total_nr_birds_hatched=n_distinct(BirdID)) %>% 
  pull(total_nr_birds_hatched) # total number of birds that hatched

#Fledged
st_2_a=piedbabbler  %>% filter(SurvFledge==1) %>% 
  summarise(total_nr_birds_fledge=n_distinct(BirdID)) %>% 
  pull(total_nr_birds_fledge) # total number of birds fledged

# Died b4 fledging
st_2_b=piedbabbler  %>% filter(SurvFledge==0) %>% 
  summarise(total_nr_birds_fledge=n_distinct(BirdID)) %>% 
  pull(total_nr_birds_fledge)



#Fledged and Surv_In
st_3_a=piedbabbler %>% filter(SurvInd==1) %>% 
  summarise(total_nr_birds_ind=n_distinct(BirdID)) %>% 
  pull(total_nr_birds_ind)


# Fledged, and Died

st_3_b=piedbabbler %>% filter(SurvInd==0) %>% 
  summarise(total_nr_birds_ind=n_distinct(BirdID)) %>% 
  pull(total_nr_birds_ind)

# Fledged, Surv_Ind and Surv_1yr
st_4_a=piedbabbler %>% filter(Surv1Yr==1) %>% 
  summarise(total_nr_birds_yr=n_distinct(BirdID)) %>% 
  pull(total_nr_birds_yr)

# Fledged, Surv_Ind and Died
st_4_b=piedbabbler %>% filter(Surv1Yr==0) %>% 
  summarise(total_nr_birds_yr=n_distinct(BirdID)) %>% 
  pull(total_nr_birds_yr)

df_sankey=data.frame(
  st_1=c('All',     'All',        'All',         'All'),
  st_2=c('Fledge',  'Fledge',     'Fledge',     'No_Fledge'),
  st_3=c('Nutr_Ind','Nutr_Ind',    'No_Nutr_Ind',  NA),
  st_4=c('Surv_1Yr','No_Surv_1Yr',  NA,            NA),
  values=c(st_4_a,st_4_b,st_3_b,st_2_b)
  
  
)

```

```{r}
library(networkD3)
library(htmlwidgets)
library(htmltools)

nodes <- data.frame(name = c("ALL",'Fledge','No Fledge','Nutr_Ind','No_Nutr_Ind','Surv_1Yr','No_Surv_1Yr'))
links=data.frame(
  source=c(0,0,1,1,3,3),
  target=c(1,2,3,4,5,6),
  value=c(st_2_a,st_2_b,st_3_a,st_3_b,st_4_a,st_4_b)
  
)

sankey=sankeyNetwork(
  Links = links,
  Nodes = nodes,
  Source = "source",      # Column name for source in 'links'
  Target = "target",      # Column name for target in 'links'
  Value = "value",        # Column name for flow value in 'links'
  NodeID = "name",        # Column name for node names in 'nodes'
  fontSize = 10,          # Adjust font size
  nodeWidth = 12          # Adjust node width
)


final_sankey=prependContent(sankey, tags$h1("Nestling Survival Flow Sankey Plot"))


final_sankey  <- htmlwidgets::onRender(
  final_sankey ,
  '
  function(el, x) {
    // select all links
    d3.select(el).selectAll(".link")
      .each(function(d) {
        // append text to parent of each link
        d3.select(this.parentNode)
          .append("text")
          .attr("x", (d.source.x1 + d.target.x0) / 2)   // middle of ribbon horizontally
          .attr("y", (d.y0 + d.y1) / 2)                 // middle of ribbon vertically
          .attr("text-anchor", "middle")
          .attr("font-size", "10px")
          .attr("fill", "black")
          .text(d.value);  // <--- This is where the value from links$value is used
      });
  }
  '
)

# Display the plot
final_sankey 
# saveWidget(final_sankey, "sankey.html", selfcontained = TRUE)
# webshot("sankey.html", "sankey.png")
```

```{r}
library(ggsankey)
library(dplyr)
library(ggplot2)

# create dataframe for each link
df <- data.frame(
  stage1 = c("All","All","Fledge","Fledge","Nutr_Ind","Nutr_Ind"),
  stage2 = c("Fledge","No Fledge","Nutr_Ind","No_Nutr_Ind","Surv_1Yr","No_Surv_1Yr"),
  value  = links$value
)

# Fill intermediate stages so make_long can work
df$stage0 <- "All"   # optional for ggalluvial/ggsankey

# Convert to long format
df_long <- df %>%
  make_long(stage1, stage2, value = value)

ggplot(df_long,
       aes(x = x, next_x = next_x,
           node = node, next_node = next_node,
           value = value)) +
  
  geom_sankey(aes(fill = interaction(node,next_node)), alpha = 0.8) +
  
  geom_sankey_label(aes(label = node), size = 4, color = "black") +
  
  # scale_x_continuous(
  #   breaks = 1:2, 
  #   labels = c("All", "Next Stage")
  # ) +
  
  theme_minimal() +
  labs(title = "Nestling Survival Flow Sankey Plot")
```

```{r}
library(ggplot2)
library(ggalluvial)

df_sankey %>%  ggplot(
       aes(axis1 = st_1,
           axis2 = st_2,
           axis3 = st_3,
           axis4 = st_4,
           y = values)) +
  
  geom_alluvium(aes(fill = path), width = 0.25) +
  
  geom_stratum(width = 0.3, fill = "grey85", color = "black") +
  
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum)),
            size = 4) +
  
  scale_x_discrete(
    limits = c("All","Fledge","Nutr_Ind","Surv_Yr"),
    expand = c(.05,.05)
  ) +
  
  labs(
    title = "Nestsling Survival Sankey",
    y = "Count"
  ) +
  
  theme_minimal()
```

```{r}
nest_all %>% filter(is.na(HatchDate)) %>% 
  summarize(n_distinct(BirdID))

fledg_survival %>% summarize(n_distinct(BirdID))
```
